### 2018-02-07 submission/manuscript.tex
change to OR

\textbf{Results.} Fecal samples had a significant decrease for both
Shannon diversity and evenness, after correcting for study effect and
variable region sequenced, with more severe disease (P-value \textless{}
0.05). This reduction in evenness translated into small increases in
relative risk for adenoma (P-value = 0.032) and carcinoma stages of CRC
(P-value = 0.00034) while the reduction in Shannon diversity only
translated into an increased relative risk for developing carcinomas
(P-value = 0.0047). Increases in mouth-associated microbes were commonly
in the top 5 most significantly increased relative risk of adenoma and
carcinoma for both stool and tissue samples. A prediction model for
adenoma and carcinoma was built using either the whole community or
selected genera with highest and lowest relative risk from fecal and
tissue samples. Both approaches resulted in similar classification
success according to Area Under the Curve (AUC) regardless of whether
genera or OTUs were used to build the model. The most important groups
within the full community models consistently belonged to genera such as
\emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia} across
studies. Although a number of associations between the microbiota and
CRC were identified, the majority of studies that we used in this
meta-analysis were only individually adequately powered for large effect
sizes.

@@

Using both feces (n = 1737) and colon tissues (492 samples from 350
individuals) totaling over 2229 total samples across 14 studies
{[}12--18,21,23--27,30{]} {[}Table 1 \& 2{]}, we expand both the breadth
and scope of the previous meta-analysis to investigate whether the
bacterial community or specific members are more important risk factors
for both adenoma and carcinoma stages of disease. To accomplish this we
first assessed whether bacterial diversity changes throughout disease
(control to adenoma to carcinoma) and if it results in an increased
relative risk (RR) for adenoma or carcinoma stages of disease We then
assessed what genera, if any, increase or decrease the RR of adenoma or
carcinoma stages of disease Next, using Random Forest models, we
analyzed whether the full community or only the combined top 5 increased
and top 5 decreased RR genera resulted in better model classification,
based on the area under the curve (AUC). Finally, we also examined at
what effect and sample size the studies used were powered for and the
sample size needed to get to the traditionally accepted 80\% power. Our
results from these analyses suggests that the bacterial community
changes as disease severity worsens, that specific members are important
for disease classification, and that many of the studies are
underpowered for assessing small effect sizes.

@@

\textbf{\emph{Lower Bacterial Diversity is Associated with Increased RR
of Tumors:}} To assess differences in broad scale community metrics as
disease severity worsens Operational Taxonomic Unit (OTU) richness,
evenness, and Shannon diversity measurements were power transformed and
Z-score normalized. These metrics are commonly used to assess the total
number of OTUs, the equality of their abundance, and the overall
diversity, respectively. Using linear mixed-effect models to control for
study and variable region we assessed whether OTU richness, evenness, or
Shannon diversity changed in a step-wise manner with disease severity.
In stool, there was a significant decrease in both evenness and Shannon
diversity as disease severity moved from control to adenoma to carcinoma
(P-value = 0.025 and 0.043, respectively) {[}Figure 1A{]}. We next
tested whether the detectable differences in community significantly
increased in RR of having an adenoma or carcinoma. For fecal samples, a
decrease versus the overall median in evenness resulted in a
significantly increased RR for carcinoma (RR = 1.36 (1.15 - 1.61),
P-value = 0.00034) and adenoma (RR = 1.16 (1.01 - 1.34), P-value =
0.032) while a decrease versus the overall median in Shannon diversity
only increased the RR for carcinoma (RR = 1.33 (1.09 - 1.62), P-value =
0.0047) {[}Figure 2{]}. Using the Bray-Curtis distance metric and
PERMANOVA, it was also possible to identify significant bacterial
community changes, in specific studies, for both carcinoma-associated
and adenoma-associated microbiota versus control {[}Table S1 \& S2{]}.

@@

\textbf{\emph{Mouth-Associated Genera are Associated with an Increased
RR of Tumor:}} Next, we asked if being higher than the median relative
abundance, for any specific genera, resulted in an altered RR for
adenoma or carcinoma, in stool and colon tissue, due to our previous
observations of small increases in RR using OTU richness and Shannon
diversity. To investigate this we analyzed all common genera across each
study, in colon tissue or stool, and assessed whether a relative
abundance higher than the median results in an increase or decrease in
RR. Mouth-associated genera were commonly found in the top 5 genera
associated with an increased RR of having an adenoma
(\emph{Pyramidobacter} {[}Figure 3A{]} and \emph{Rothia} {[}Figure
3C{]}) and carcinoma (\emph{Fusobacterium}, \emph{Parvimonas},
\emph{Porphyromonas}, and \emph{Peptostreptococcus} {[}Figure 3B{]} and
\emph{Fusobacterium} {[}Figure 3D{]}) for both stool and colon tissue
samples. Conversely, genera commonly associated with a normal
gastrointestinal tract were correlated with a decreased RR for both
adenoma and carcinoma for both stool and colon tissue samples {[}Figure
3{]}. Even though mouth-associated genera were identified across disease
stage, there was little direct overlap of the top 5 increased or
decreased RR genera between both stages and sample site.

When observing RRs with a P-value less than 0.05 there was almost no
overlap between genera from stool or colon tissue and when they were
similar the RR was in opposite directions (e.g. \emph{Lactococcus})
{[}Table S8 \& S9{]}. Many of the genera that had RRs with a P-value
under 0.05 for colon tissue are also highly prevalent in contamination,
specifically, \emph{Novosphingobium}, \emph{Selemonas}, and
\emph{Achromobacter} {[}Figure 3 \& Table S8-S9{]}. For carcinoma stage
of disease, certain mouth-associated genera (\emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}) had a high RR for both
colon tissue and stool samples {[}Table S10 \& S11{]}. Finally, these
data suggest that the most significantly increased RR genera for tissue
was \emph{Camplyobacter} while in stool it was \emph{Peptostreptococcus}
{[}Table S10 \& S11{]}.

\textbf{\emph{Select Community Models can Recapitulate Whole Community
Models:}} Since specific genera increased RR for carcinoma over
diversity metrics we assessed whether the bacterial community was better
at classifying disease versus only a select group of genera. We selected
these genera based on their RR and P-value significance and used two
approaches to test this question. The first approach used genus level
data and tested for differences in AUC between all genera and selected
genera. A single study was used for training the model prior to testing
on all other studies and this was repeated for every study in the
meta-analysis. The second approach used OTU level data and tested for a
generalized decrease in the 10-fold cross validation (CV) model AUC
which is a common approach used to guard against over-fitting. This was
applied across study and the AUC of the all OTUs model was compared
against the model that used only OTUs that taxonomically classified to
selected genera.

@@

Our study identifies clear differences in diversity, both at the
community level and for individual genera, present in patients with and
without CRC {[}Figure 1-3{]}. Although there was a step-wise decrease in
diversity as disease progressed from control to adenoma to carcinoma,
this did not translate into large effect sizes for the RR of tumors.
Even though mouth-associated genera increased the RR of having a
carcinoma, they did not consistently increase the RR of having an
adenoma. Additionally, our observations suggest that by combining
mouth-associated and CRC protective microbes we can classify either
adenoma or carcinoma stage of disease as well as models that use the
full community.

@@

\textbf{\emph{Obtaining Genera Relative Abundance and Selected Models:}}
For the genera analysis of the RR, OTUs were added together based on the
genus or lowest available taxonomic classification level and the total
average counts, for 100 different subsamplings, were collected. The OTU
based Random Forest Models using selected OTUs utilized a similar
approach except that the OTUs were not aggregated together by taxonomic
identity but kept as separate OTUs. OTU Random Forest models using the
full community included all OTUs while those for the selected model
included only those OTUs that had a taxonomic classification to a
variable in the top 5 increased of top 5 decreased RR based on P-value.

>>>>

\textbf{Results.} Fecal samples had a significant decrease for both
Shannon diversity and evenness, after correcting for study effect and
variable region sequenced, with more severe disease (P-value \textless{}
0.05). This reduction in evenness translated into small increases in the
odds ratio for adenoma (P-value = 0.035) and carcinoma stages of CRC
(P-value = 0.0021) while the reduction in Shannon diversity only
translated into an increased odds ratio for developing carcinomas
(P-value = 0.0069). Increases in mouth-associated microbes were commonly
in the top 5 most significantly increased relative risk of adenoma and
carcinoma for both stool and tissue samples. A prediction model for
adenoma and carcinoma was built using either the whole community or
selected genera with highest and lowest relative risk from fecal and
tissue samples. Both approaches resulted in similar classification
success according to Area Under the Curve (AUC) regardless of whether
genera or OTUs were used to build the model. The most important groups
within the full community models consistently belonged to genera such as
\emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia} across
studies. Although a number of associations between the microbiota and
CRC were identified, the majority of studies that we used in this
meta-analysis were only individually adequately powered for large effect
sizes.

@@

Using both feces (n = 1737) and colon tissues (492 samples from 350
individuals) totaling over 2229 total samples across 14 studies
{[}12--18,21,23--27,30{]} {[}Table 1 \& 2{]}, we expand both the breadth
and scope of the previous meta-analysis to investigate whether the
bacterial community or specific members are more important risk factors
for both adenoma and carcinoma stages of disease. To accomplish this we
first assessed whether bacterial diversity changes throughout disease
(control to adenoma to carcinoma) and if it results in an increased odds
ratio (OR) for adenoma or carcinoma stages of disease We then assessed
what genera, if any, increase or decrease the OR of adenoma or carcinoma
stages of disease Next, using Random Forest models, we analyzed whether
the full community or only the combined top 5 increased and top 5
decreased OR genera resulted in better model classification, based on
the area under the curve (AUC). Finally, we also examined at what effect
and sample size the studies used were powered for and the sample size
needed to get to the traditionally accepted 80\% power. Our results from
these analyses suggests that the bacterial community changes as disease
severity worsens, that specific members are important for disease
classification, and that many of the studies are underpowered for
assessing small effect sizes.

@@

\textbf{\emph{Lower Bacterial Diversity is Associated with Increased OR
of Tumors:}} To assess differences in broad scale community metrics as
disease severity worsens Operational Taxonomic Unit (OTU) richness,
evenness, and Shannon diversity measurements were power transformed and
Z-score normalized. These metrics are commonly used to assess the total
number of OTUs, the equality of their abundance, and the overall
diversity, respectively. Using linear mixed-effect models to control for
study and variable region we assessed whether OTU richness, evenness, or
Shannon diversity changed in a step-wise manner with disease severity.
In stool, there was a significant decrease in both evenness and Shannon
diversity as disease severity moved from control to adenoma to carcinoma
(P-value = 0.025 and 0.043, respectively) {[}Figure 1A{]}. We next
tested whether the detectable differences in community significantly
increased in OR of having an adenoma or carcinoma. For fecal samples, a
decrease versus the overall median in evenness resulted in a
significantly increased RR for carcinoma (OR = 1.66 (1.2 - 2.3), P-value
= 0.0021) and adenoma (OR = 1.3 (1.02 - 1.65), P-value = 0.035) while a
decrease versus the overall median in Shannon diversity only increased
the OR for carcinoma (OR = 1.61 (1.14 - 2.28), P-value = 0.0069)
{[}Figure 2{]}. Using the Bray-Curtis distance metric and PERMANOVA, it
was also possible to identify significant bacterial community changes,
in specific studies, for both carcinoma-associated and
adenoma-associated microbiota versus control {[}Table S1 \& S2{]}.

@@

\textbf{\emph{Mouth-Associated Genera are Associated with an Increased
OR of Tumor:}} Next, we asked if being higher than the median relative
abundance, for any specific genera, resulted in an altered OR for
adenoma or carcinoma, in stool and colon tissue, due to our previous
observations of small increases in OR using OTU richness and Shannon
diversity. To investigate this we analyzed all common genera across each
study, in colon tissue or stool, and assessed whether a relative
abundance higher than the median results in an increase or decrease in
RR. Mouth-associated genera were commonly found in the top 5 genera
associated with an increased OR of having an adenoma
(\emph{Pyramidobacter} {[}Figure 3A{]} and \emph{Rothia} {[}Figure
3C{]}) and carcinoma (\emph{Fusobacterium}, \emph{Parvimonas},
\emph{Porphyromonas}, and \emph{Peptostreptococcus} {[}Figure 3B{]} and
\emph{Fusobacterium} {[}Figure 3D{]}) for both stool and colon tissue
samples. Conversely, genera commonly associated with a normal
gastrointestinal tract were correlated with a decreased OR for both
adenoma and carcinoma for both stool and colon tissue samples {[}Figure
3{]}. Even though mouth-associated genera were identified across disease
stage, there was little direct overlap of the top 5 increased or
decreased OR genera between both stages and sample site.

When observing ORs with a P-value less than 0.05 there was almost no
overlap between genera from stool or colon tissue and when they were
similar the RR was in opposite directions (e.g. \emph{Lactococcus})
{[}Table S8 \& S9{]}. Many of the genera that had ORs with a P-value
under 0.05 for colon tissue are also highly prevalent in contamination,
specifically, \emph{Novosphingobium}, \emph{Selemonas}, and
\emph{Achromobacter} {[}Figure 3 \& Table S8-S9{]}. For carcinoma stage
of disease, certain mouth-associated genera (\emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}) had a high OR for both
colon tissue and stool samples {[}Table S10 \& S11{]}. Finally, these
data suggest that the most significantly increased RR genera for tissue
was \emph{Camplyobacter} while in stool it was \emph{Peptostreptococcus}
{[}Table S10 \& S11{]}.

\textbf{\emph{Select Community Models can Recapitulate Whole Community
Models:}} Since specific genera increased OR for carcinoma over
diversity metrics we assessed whether the bacterial community was better
at classifying disease versus only a select group of genera. We selected
these genera based on their OR and P-value significance and used two
approaches to test this question. The first approach used genus level
data and tested for differences in AUC between all genera and selected
genera. A single study was used for training the model prior to testing
on all other studies and this was repeated for every study in the
meta-analysis. The second approach used OTU level data and tested for a
generalized decrease in the 10-fold cross validation (CV) model AUC
which is a common approach used to guard against over-fitting. This was
applied across study and the AUC of the all OTUs model was compared
against the model that used only OTUs that taxonomically classified to
selected genera.

@@

Our study identifies clear differences in diversity, both at the
community level and for individual genera, present in patients with and
without CRC {[}Figure 1-3{]}. Although there was a step-wise decrease in
diversity as disease progressed from control to adenoma to carcinoma,
this did not translate into large effect sizes for the OR of tumors.
Even though mouth-associated genera increased the OR of having a
carcinoma, they did not consistently increase the RR of having an
adenoma. Additionally, our observations suggest that by combining
mouth-associated and CRC protective microbes we can classify either
adenoma or carcinoma stage of disease as well as models that use the
full community.

@@

\textbf{\emph{Obtaining Genera Relative Abundance and Selected Models:}}
For the genera analysis of the RR, OTUs were added together based on the
genus or lowest available taxonomic classification level and the total
average counts, for 100 different subsamplings, were collected. The OTU
based Random Forest Models using selected OTUs utilized a similar
approach except that the OTUs were not aggregated together by taxonomic
identity but kept as separate OTUs. OTU Random Forest models using the
full community included all OTUs while those for the selected model
included only those OTUs that had a taxonomic classification to a
variable in the top 5 increased of top 5 decreased OR based on P-value.
